---
title: "Data Frames"
author: "Ahmad Abdalla"
date: "11/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## DataFrame
###  Data frame is a two dimensional data structure in R. It is a special case of a list which has each component of equal length.
#### • Data frames are used to store tabular data in R.
#### • It is an Important type of object in R, used in a variety of statistical modeling applications.
#### • Unlike matrices, data frames can store different classes of objects in each column.
#### • Data frames have a special attribute called row.names which indicate information about each row of the data frame.

#### • Data frames are usually created by reading in a dataset using the read.table(..) or read.csv(..).
#### • Data frames can also be created explicitly with the data.frame(..) function or they can be coerced from other types of objects like lists.

### Creating Data Frames from Vectors
```{r}
n = c(2 , 3, 5)
s = c("aa", "bb", "cc")
b = c(TRUE , FALSE , TRUE )
df = data.frame (n, s, b) # df is a data frame
df
```


### downloading mtcars data frame

```{r cars}
data(mtcars)
mtcars
```

### checking out the meta data for mtcars using these functions

```{r pressure, echo=FALSE}
class(mtcars)
typeof(mtcars)
str(mtcars)
summary(mtcars)
row.names(mtcars)
head(mtcars)
tail(mtcars)
View(mtcars)
ncol(mtcars)
nrow(mtcars)
```
### how to retrieve data value from row 1, column 2
```{r}
mtcars[1,2]
```

### how to retrieve data value from row 1, column 2 using the names
```{r}
mtcars["Mazda RX4", "cyl"]

```
### how to retrieve data of a row eg Volvo 142E

```{r}
mtcars["Volvo 142E",]
```
### retrieve two rows (any rows, up to you)
```{r}
mtcars[5:6,]
```
### retrieve data for column mpg
```{r}
mtcars[,"mpg"]
```
### what is the class of the output given to you by R?
```{r}
class(mtcars[,"mpg"])
```
### is it a data frame?
### how do you get back data frame

```{r}
class(mtcars[,"mpg", drop = FALSE])
```
### can you use drop when retrieving data using names ?

```{r}
print("yes you can as been tried above")
```

## Factors
#### • Factor is a data structure used for fields that takes only predefined, finite number of values (categorical data). For example: a data field such as marital status may contain only values from single, married, separated, divorced, or widowed.
#### • In such case, we know the possible values beforehand and these predefined, distinct values are called levels.

#### ►Factors are used to represent categorical data.
#### ►One can think of a factor as an integer vector where each integer has a label.
#### ►Factors are important in statistical modeling.
#### ►Using factors with labels is better than using integers because factors are self-describing.

### to create a factor object we use the factor() function
```{r}
x<-factor(c("single","married","married","single"));
x
```
### the levels of a factor maybe defined even if not used:
```{r}
x<-factor(c("single","married","married","single"),
levels=c("single","married","divorced"));
x
str(x)

```
### Levels are stored in a character vector and the individual elements are actually stored as indices. Accessing components of a factor is very similar to that of vectors.
### accessing factors
#### x[3] #access the third element
#### x[4]
#### x[c(2,4)] #access the 2nd and 4th
#### x[-1] #access all but 1st
#### x[c(TRUE,FALSE,FALSE,TRUE)]#uselogical vector for 1st and last

### Invalid assignment
```{r}
x[3]<-"widowed" #cannot assign outside values
```
### A workaround to this is to add the value to the level first. 
```{r}
levels(x)<-c(levels(x), "widowed") #add widowed to level
x
x[3]<-"widowed" #widowed already added to level
 x
```