---
title: "Control Structure"
author: "Ahmad Abdalla"
date: "11/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Control statements
#### • if statements
#### • The language has available a conditional construction of the form if (expr 1) expr 2 else expr 3
#### • where expr 1 must evaluate to a logical value and the result of the entire expression is then evident.
#### • a vectorized version of the if/else construct, the ifelse function. This has the form ifelse(condition, a, b)

### • for loops, repeat and while
#### • for (name in expr 1) expr 2 where name is the loop variable. expr 1 is a vector expression, (often a sequence like 1:20), and expr 2 is often a grouped expression with its sub-expressions written in terms of the dummy name. expr 2 is repeatedly evaluated as name ranges through the values in the vector result of expr 1.Control statements

### If statement
#### here is an example:
```{r}
voter <- T
age <- 10
if(age <18){
  message("this is a minor")
  voter <- F
}
message("if statement is completed")
voter
```
### If elese statement
#### here is an example:
```{r}
voter <- T
age <- 20
if(age <18){
  message("this is a minor")
  voter <- F
} else{
  message("this is an adult")
  voter <- T
}
voter
```
### IFElse statement, quick assignment for a vector
```{r}
age <- c(12, 18, 22, 3)
names(age) <- c("omar", "ibrahim", "ahmed", "salah")
voter <- ifelse(age<18, "minor", "voter")
voter
```
### if statement with else if statement
```{r}
voter <- T
drinker <- T
age <- 20
  if(age <18){
  message("this is a minor")
    drinker <- F
  voter <- F} else if(age <21){
  message("this is a voter")
  voter <- T
  drinker <- F
      } else{
  message("this is an adult")
  voter <- T
  drinker <- T
        }

drinker
voter

```
### Switch
#### • serves like the ifelse function for multiple conditions
```{r}
dish <- "fruit"
switch(dish, fruit = "banana", vegetable = "broccoli", meat = "chicken")
x <- switch (3,"first", "second", "third", "fourth")
x
```
### loops
#### • When the same or similar tasks need to be performed multiple times; for all elements of a list; for all columns of an array; etc.
####  • Monte Carlo Simulation
#### • Cross-Validation (delete one and etc)
### example 1 for a loop using for statement:
```{r}
for(i in 1:10){
  print(i*i)
}
```
### example 2 for a loop using while statement:
```{r}
i=1
while (i <=10) {
  print(i*i)
  i = i +sqrt(i)
  
}
```
### For loops repeat the same task for changing values of a given variable.
```{r}
ages <- c(12, 16, 19, 25, 35)
names(ages) <- c("abdalla", "omar", "nawaf", "yousif", "meshari")
for (person in names(ages)) {
  message(person, " is ", ages[person], " years old")
  
}
```
###  Can set variable values within a loop.
```{r}
SumAge <- 0
ages <- c(12, 16, 19, 25, 35)
names(ages) <- c("abdalla", "omar", "nawaf", "yousif", "meshari")
for (person in names(ages)) {
  message(person, " is ", ages[person], " years old")
  SumAge = SumAge + ages[[person]]
}
SumAge
```
### Nested Loops
```{r}
ages <- c(12, 16, 23, 35, 50)
names(ages) <- c("abdalla", "omar", "nawaf", "yousif", "meshari")
for (person in names(ages)) {
 message(person)
  for (i in 1:ages[person]) {
    message("year ", i)
  }
}
```
### Embedding if statements in for loops
```{r}
ages <- c(12, 16, 23, 35, 50)
names(ages) <- c("abdalla", "omar", "nawaf", "yousif", "meshari")
numVoters <- 0
for (person in names(ages)) {
  
  if( ages[person]>=18){
    message(person, " can vote")
    numVoters = numVoters +1
  }
}
numVoters
```
### Setting values with embedded if statements in for loops
```{r}
numVoters <- 0
ages <- c(12, 16, 23, 35, 50)
names(ages) <- c("abdalla", "omar", "nawaf", "yousif", "meshari")
voters <- vector()
for (person in names(ages)) {
  if(ages[person] >= 18){
    message(person, " can vote")
    numVoters <- numVoters +1
    voters[numVoters] <- person
  }
}
numVoters
voters
```
### While Loop
#### while loops repeat the same operation until a test condition is met.
##### • Note: Be very careful with while loops that the condition can become FALSE or you can be trapped in an infinite loop!!!!
```{r}
z <- 0
while (z<5) {
  z = z+1
  message(z)
  
}
```
### repeat loops
#### • A repeat loop is used to iterate over a block of code multiple number of times. (There is no condition)
```{r}
x <- 1
repeat{
  print(x)
  x = x +1
  if(x == 5){
    break
  }
}

```
### Break Statement
#### • A break statement is used inside a loop (repeat, for, while) to stop the iterations and flow the control outside of the loop.
#### • In a nested looping situation, where there is a loop inside another loop, this statement exits from the innermost loop that is being evaluated.
```{r}
x <- 1:10
for (val in x) {
  if(val == 7){
    break
  }
  print(val)
}
```
### Next statement 
#### next statement is useful when we want to skip the current iteration of a loop without terminating it. On encountering next, the R parser skips further evaluation and starts next iteration of the loop.
```{r}
x <- 1:10
for (val in x) {
  if(val == 4){
    next
  }
  print(val)
}
```