---
title: "Dplyer package"
author: "Ahmad Abdalla"
date: "12/8/2021"
output: html_document
---


### Loading Dplyer library after installing it 
```{r}
library(dplyr)
```
### One handy feature with dplyr is the glimpse(..) function. This is  an improvement over str(..).
#### Use glimpse(..) to see the structure of the dataset and  decide  what manipulation is required.
```{r}
glimpse(mtcars)
```
### sum()
#### Count the observations in a column in the data frame.
```{r}
sum(mtcars$cyl ==8)
sum(mtcars$gear > 4 )
```
### select()
#### Selects a range of variables and can exclude some variables

#### select(df,A,B,C)	Select the variables A,B and C  from df dataset.
#### select(df,A:C)		Select all variables from A to C from df dataset.
#### select(df, -C)		Exclude C from the dataset from df  dataset.
#### Can use variable names as well as integer index

```{r}
select(mtcars, cyl:wt)
select(mtcars, cyl:wt, -hp)
select(mtcars, 1:4, -3)

select (mtcars, starts_with("c"))
select (mtcars, starts_with("c"),   ends_with("t"))

```
### Helper Functions
#### Dplyr has helper functions to help select a group of variables inside a select()
#### starts_with(“x”) 	every name that starts with “x”
#### ends_with(“x”)	every name that ends with “x”
#### contains(“x”)	every name that contains “x”
#### matches(“x”)	every name that matches “x”, where “x” can be a regular expression
#### num_range(“x”, 1:5)	the variable named x01, x02, x03, x04, x05
#### one_of(x)		every name that appears in x, which should be a character vector

#### Note : when referring to variables directly inside select(), don’t use quotes “”. When using helper function, use quotes
```{r}
select(mtcars, contains('c')) 
select(mtcars, matches('a')) 
select(mtcars, num_range("mpg", 1:5)) 
select(mtcars, one_of("mpg", "cyl"))
```
### mutate(
#### Creates new column and add to a copy of the dataset
#### Syntax:   
#### mutate(df, z=x+y)
#### Can add multiple variables
#### Can also use the newly created variable
#### mutate(df, z=x+y, zz = z+v)
```{r}
mtM <- mutate(mtcars, hpp = hp/10)
mtM
```
### filter()
#### The filter(..) helps to keep the observations following a criteria.
#### The filter(..) works exactly like select(), you pass the data  frame first and then a condition separated by a comma
#### filter(df, condition)
#### We can filter a dataset with more than one criteria.
```{r}
filter(mtcars, cyl == 8, qsec >16)
filter(mtcars, cyl == 8 & qsec >16)
filter(mtcars, (cyl == 8 | cyl == 6) & qsec >16)
filter(mtcars, cyl %in% c(4,8))
filter(mtcars, !is.na(mpg))  # choose only observations which value of x is not NA
```
### arrange()
#### To arrange rows according to any type of data
#### Default is ascending. To arrange in descending order, use desc()
```{r}
arrange(mtcars, cyl)
arrange(mtcars, cyl, mpg)
arrange(mtcars, cyl, desc(mpg))
```
### arranging a filtered data set
```{r}
arrange(filter(mtcars, (cyl == 8 | cyl == 6) & qsec >16), desc(cyl))
```
### arranging a filtered data set with selected variables
```{r}
arrange(filter(select(mtcars, mpg:vs, -disp, -drat), (cyl == 8 | cyl == 6) & qsec >16), desc(cyl) )
```
### summarise()
#### summarise() creates a new data frame. It will have one (or more) rows for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarising all observations in the input. It will contain one column for each grouping variable and one column for each of the summary statistics that you have specified.
```{r}
summarise(mtcars, mean = mean(mpg), n = n())
summarise(mtcars, min_mpg = min(mpg))
summarise(mtcars, max_mpg = max(mpg),min_mpg = min(mpg))
```
#### In summarise(), can use any aggregating function that can take a vector of data and return a single number including min(), max(), mean(), median(), quantile(), sd(), var()
```{r}
summarise(filter(mtcars, cyl ==8), min_mpg = min(mpg), max_mpg = max(mpg))
```
### pipeline
#### The dplyr library comes with a practical operator, % >%, called the  pipeline. The pipeline feature makes the manipulation clean, fast and less  prompt to error. use the pipeline operator %>%  define  the data frame used at the beginning and all the process will flow  from it.
```{r}
newmtpipe <- mtcars %>% select ( mpg:vs, -disp, -drat) %>%
  filter((cyl == 8 | cyl == 6) & qsec >16) %>% arrange(desc(cyl))
newmtpipe
```
#### another example for a pipeline
```{r}
newsumpipe <- mtcars %>% filter(cyl == 8) %>% summarise(min_mpg = min(mpg), max_mpg = max(mpg))
newsumpipe
```


