---
title: "Data Wrangling - Cleaning your Data (tidyr)"
author: "Ahmad Abdalla"
date: "12/15/2021"
output: html_document
---

## TIBBLE
#### Whenever working with rectangular data structures — data consisting of multiple cases (rows) and variables (columns) — our first step (in a tidyverse context) is to create or transform the data into a tibble. A tibble is a rectangular data table and a modern and simpler version of the data.frame construct in R.

### Loading tidyvesre library
```{r}
library(tidyverse)
```
#### tibbles do not change the types of input variables (e.g., strings are not converted to factors);
#### tibbles do not change the names of input variables and do not use row names.
#### Tibbles are quite flexible in also allowing for non-syntactic variable (column) names. For instance, in contrast to data frames, the variable names in tibbles can start with a number or contain spaces

### changing mtcars data frame to tibble
```{r}
mtc<-as_tibble(mtcars)
class(mtc)
mtc
```
### runing some of the dplyr codes to see how they interact with mtc which is a tibble now
```{r}
sum(mtc$gear > 4 )
select(mtc, 1:4, -3)
select (mtc, starts_with("c"),   ends_with("t"))
select(mtc, num_range("mpg", 1:5))
mtM <- mutate(mtc, hpp = hp/10)
mtM
filter(mtc, cyl %in% c(4,8))
arrange(mtc, cyl, desc(mpg))
arrange(filter(select(mtc, mpg:vs, -disp, -drat), (cyl == 8 | cyl == 6) & qsec >16), desc(cyl) )
summarise(mtc,  max_mpg = max(mpg),min_mpg = min(mpg))
newmtpipe <- mtc %>% select ( mpg:vs, -disp, -drat) %>%
  filter((cyl == 8 | cyl == 6) & qsec >16) %>% arrange(desc(cyl))
newmtpipe
```
## TIDYR
### it is a package that reshape the layout of tables
### Gather()
#### It collapses multiple columns into key-variable pair
#### the input of gather are: gather(df, key, value, column), where key is the column that represents the variables that are converted to columns, and value represent the column that has the values from the previous columns.
```{r}
messy <- data.frame("name" = c("ahmed", "khalid", "yousif"), a = c(50, 80, 90), b = c(70, 90, 10))
messy
gather(messy, drug, heartrate, 2:3)
```
### spread()
#### • takes two columns (key & value) and spreads in to multiple columns
#### • Convert long data to wide. spread(df, key, value)
```{r}
area <- data.frame(city = c("madinah", "jeddeh", "ryadh","madinah", "jeddeh", "ryadh"),
                   size = c("small", "large","small","large", "small","large"),
                   amount= c(40,100,120,50, 90, 110))
area
spread(area, size, amount)
```
### separate()
#### • Multiple variables are stored in one column .
#### • Separate into multiple columns
#### • separate(df, data, colname)
```{r}
sep <- data.frame(a = c("1.a", "2.b", "3.c"))
sep
seped <- separate(sep, a, c("num", "char"))
seped
```
### unite()
#### • To join variables 
#### • unite(df, newcol, original_colums)
```{r}
united <- unite(seped, a, c("num", "char"))
united
```
## lubridate
### a package that helps date conversion
#### loading the lubridate package
```{r}
library(lubridate)
```
### experimenting with basic lubridate functions
```{r}
ymd("2015-8-25")
ymd("2015 August 25")
mdy("August 25, 2015")
hms("08:44:34")
ymd_hms("2015/08/25 08.44.34")
```
## stringr
### A package that works with strings
#### loading the stringr package
```{r}
library(stringr)
```
### experimenting with basic stringr functions
```{r}
str_trim("  this this is ahemd   ")
str_pad("0551", width = 7, side = "left", pad = "5")
friends <- c("salah", "ibrahim", "omar", "osama")
str_detect(friends, "ibrahim")
str_replace(friends, "omar", "odai")

```
### Other cleaning processes:
#### • toupper(), Change string touppercases
#### • tolower(), Change string touppercases
```{r}
upper <- toupper("hi I'm funny")
upper
lower <- tolower(upper)
lower
```
## missing values
### Use is.na() to find NA values in dataset
```{r}
x <- c(1,2,3,NA)
is.na(x)
!is.na(x)
```
#### Exclude missing value from analyses
```{r}
x <- c(1,2,NA,3)
mean(x) # returns NA
mean(x, na.rm=TRUE) 
complete.cases(x)
any(is.na(x)) #use any to check if there is any NA values
na.omit(x) # to remove the NA values, in dataframe this function remove the rows that contain NA values.
```
## visualizing some graphs
### histogram
```{r}
hist(mtcars$mpg)
```
### boxplot
```{r}
boxplot(mtcars$mpg)
```
