---
title: "Plots"
author: "Ahmad Abdalla"
date: "1/5/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic plots

```{r}
plot( mtcars$wt, mtcars$mpg) #default type p
plot(mtcars$wt, mtcars$mpg, type = "l")
plot(mtcars$wt, mtcars$mpg, type = "p")
plot(mpg ~ wt, data = mtcars)
plot(wt ~ mpg, data = mtcars)
plot(wt, mpg, data= mtcars) #error
plot(mtcars$wt, mtcars$mpg, main = "mtcars dataset", xlab = "X axis wt", ylab = "Y axis mpg")
plot(mtcars$wt, mtcars$mpg, main = "mtcars dataset", xlab = "X axis wt", ylab = "Y axis mpg",
pch = 19, frame = F)
line(lowess(mtcars$wt, mtcars$mpg), col = "blue")
```

## Understanding the ggplot Syntax
### ggplot works with dataframes and not individual vectors.
### All the data needed to make the plot is typically be contained within the dataframe supplied to the ggplot() itself or can be supplied to respective geoms.
### Another noticeable feature is that you can keep enhancing the plot by adding more layers (and themes) to an existing plot created using the ggplot() function.

## examples:
```{r}
qplot(wt, mpg, data = mtcars, geom= c("point"))
qplot(wt, mpg, data = mtcars, geom= c("smooth"))
qplot(wt, mpg, data = mtcars, geom= c("point","smooth"))
qplot(wt, mpg, data = mtcars, color=cyl) #better if char or factor
qplot(mpg, wt, data = mtcars, colour = cyl)
qplot(mpg, data = mtcars, fill=cyl ) #the data doesnt reflect
mt<-mtcars
mt$cyl<- as.factor(mt$cyl)
qplot(mpg, wt, data = mt, colour = cyl) #better with factor data
```

## ggplot2 package examples while supplying the edits to respective geoms
```{r}
ggplot(mtcars, aes(wt, mpg)) + geom_point()
ggplot(mtcars, aes(wt, mpg)) + geom_point() + geom_smooth() #default method
ggplot(mtcars, aes(wt, mpg)) + geom_point() + geom_smooth(method = "lm")
ggplot(mtcars, aes(wt, mpg)) + geom_point(aes(color=cyl))
ggplot(mtcars, aes(wt, mpg)) + geom_point(aes(color= factor(cyl))) #can change to factor
ggplot(mtcars, aes(x = factor(cyl))) + geom_bar()
```
## adding labels and titles
```{r}
g2<-ggplot(mtcars, aes(wt, mpg)) + geom_point(aes(color= factor(cyl))) #can change to factor
g2 + labs( title =" Mpg vs Wt ", subtitle =" optional title",
y=" mpg ", x=" wt ", caption =" mtcars Dataset ")
#or
g2 + ggtitle (" Mpg vs Wt ", subtitle =" optional title") +
xlab (" Mpg ") + ylab (" wt ")
#or full
ggplot(mtcars, aes(wt, mpg)) + geom_point(aes(color= factor(cyl))) +
labs( title =" Mpg vs Wt ", subtitle =" optional title",
y=" mpg ", x=" wt ", caption =" mtcars Dataset ")
```